---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# aphantasiaEmotions

<!-- badges: start -->
<a href="https://osf.io/b837s/" target="_blank"><img alt="OSF badge" src="https://img.shields.io/badge/OSF-https://osf.io/b837s/-337AB7?logo=osf"/></a>
<!-- badges: end -->

aphantasiaEmotions is a data analysis project wrapped in an R package for
reproducibility[^1]. It contains the code and data to reproduce the analyses 
presented in the article "*Aphantasics Process Emotions Differently: Evidence of 
a Non-linear Relationship Between Visual Imagery and Emotional Processing Abilities*". 
You can read the preprint <a href="https://osf.io/b837s/" target="_blank">here</a>.
All study materials are available on the Open Science Framework 
<a href="https://osf.io/b837s/" target="_blank">here</a>.

[^1]: The R package structure was chosen to facilitate the sharing of the code 
and data with the scientific community, and to make it easy to reproduce the 
analyses. It is not intended to be a general-purpose package, but rather a
collection of functions and data specific to this study (although many functions
are reusable in their own right). The package development workflow (see
<a href="https://r-pkgs.org/" target="_blank">this reference book</a>) is also a 
good way to ensure that the code is well-documented and tested, which is 
important for reproducibility in scientific research.

## Installation

You can install the development version of aphantasiaEmotions from 
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("m-delem/aphantasiaEmotions")
```

Alternatively, you can clone the repository, launch the R project in RStudio by 
opening the `aphantasiaEmotions.Rproj` file and run the following command:

```{r}
devtools::load_all()
```

... Which will load the package and make all its functions and data available in
your R session.
