---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This page is a short, practical introduction to `aphantasiaEmotions`: how to
load it, what the data looks like, and how to run one basic model. It is not
the place to learn about the study's findings; for that, see the
[Extended Online Report](https://m-delem.github.io/aphantasiaEmotions), starting
with [how this study found its shape](https://m-delem.github.io/aphantasiaEmotions/articles/how-this-study-found-its-shape.html).
This page is about *using the package*, for anyone who wants to reproduce or
build on the analyses directly in R.

```{r setup}
library(aphantasiaEmotions)
```

## The data

The package ships with one built-in dataset, `all_data`: the pooled sample
from five studies, combined and cleaned for this project's analyses.

```{r glimpse-data}
dplyr::glimpse(all_data)
```

A few columns worth knowing about from the start:

- **`study`** identifies which of the five source datasets a row comes from
  (`"burns"`, `"monzel"`, `"ruby"`, `"mas"`, `"kvamme"`) — see the
  [sample description](https://m-delem.github.io/aphantasiaEmotions/articles/sample-description.html)
  page for what each one is.
- **`vviq`** is the raw Vividness of Visual Imagery Questionnaire total score
  (16-80, where 16 is a complete absence of voluntary visual imagery).
- **`tas`** is the raw Toronto Alexithymia Scale (TAS-20) total score.
- **`vviq_group_4`** and **`vviq_group_2`** are categorical groupings of
  `vviq` at different levels of granularity, used throughout this project's
  more traditional, categorical analyses.

## Fitting a model

Most of this project's statistical work uses `fit_brms_model()`, a thin
wrapper around `brms::brm()` with this project's own conventions for chains,
iterations, and priors already set as sensible defaults. Here is the
simplest possible example, a plain linear model of alexithymia as a
function of imagery vividness:

```{r simple-model, eval = FALSE}
simple_model <- fit_brms_model(
  formula = tas ~ vviq,
  data = all_data,
  prior = brms::prior(normal(0, 20), class = "b")
)

summary(simple_model)
```

This example is not run on this page (fitting a real Bayesian model takes
real time), but it will work as written if you run it yourself. See
`?fit_brms_model` for the full set of arguments, including how to control
the number of chains and iterations.

## Where to go next

- To reproduce a specific analysis from the paper, the
  [model comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
  and [floor-group model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html)
  pages show the actual code behind each result, narrated in full.
- To see every function this package exports, browse the
  [reference index](https://m-delem.github.io/aphantasiaEmotions/reference/index.html).
- To understand *why* the study took the shape it did before diving into the
  numbers, start with
  [how this study found its shape](https://m-delem.github.io/aphantasiaEmotions/articles/how-this-study-found-its-shape.html).
