---
title: "The history of a study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The history of a study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
```

This study did not begin as a study. It began as a side question attached to a
different one — and the shape it ended up taking was not planned in advance so
much as discovered, one dataset at a time. Most papers describe what was found,
not how the finding was arrived at. This page exists to tell that second story —
the searching, the emailing, the re-analysing after every new dataset arrived.
This page is about the discovery process.

## Where it started

I was preparing for a larger planned study on aphantasia and alexithymia,
together with Gaën Plancher, Marine Mas, and Olivier Luminet. That study was
going to use a fuller battery of questionnaires than the usual VVIQ (visual
imagery) and TAS-20 (alexithymia) alone, since we were, and remain, fairly
critical of what those two standard instruments actually capture on their own.

Before designing that study, I wanted to ground it in what was already known. So
the first, modest step was simply to see how much existing data on the VVIQ and
TAS-20 already existed, to get a sense of expected effect sizes and to plan
sensible sequential analyses.

## Ale and Burns, and a locked door

On 2025-10-13, I wrote to Edwin Burns about a preprint he and his co-author had
posted the year before, on aphantasia, alexithymia, and PTSD symptomatology:
[doi.org/10.31234/osf.io/kj5d3](https://doi.org/10.31234/osf.io/kj5d3). Their
paper reported analyses of alexithymia and PTSD separately in relation to
aphantasia, but not the direct relationship between aphantasia and alexithymia
themselves — the co-occurrence question. I asked whether that had simply not
been reported, and mentioned that I'd been curious to look at their data myself,
except the OSF link in the paper pointed to a private repository.

Edwin Burns kindly shared the dataset. A short back-and-forth followed in which
we flagged what turned out to be a real discrepancy in how the TAS-20 total
score had been computed in the file — the kind of routine data-quality check
that's worth doing on any second-hand dataset before trusting it, and a good
habit to have gotten into early, since it would matter again later with a much
larger dataset. By 2025-10-29 the correction was confirmed and the data were
usable.

Ale and Burns' final sample: 192 English-speaking participants (122 females, 3
other genders; mean age 38.7, SD 11.4; range 18-86), recruited via social media,
openly archived on OSF (<https://doi.org/10.17605/OSF.IO/B837S>).

## Monzel, and the first pattern with a finer split

I collected the Monzel et al. (2024) dataset in the meantime — 105
English-speaking participants (74 females; mean age 27.9, SD 9.29; range 18-59),
recruited via the *Aphantasia Research Project Bonn*'s participant database,
also openly archived on OSF
([osf.io/y9c8g](https://osf.io/y9c8g/?view_only=1e6bd8670a3f4eacb1cf0f600343205e)).

Monzel et al.'s original paper had split their own sample at VVIQ \< 23
(aphantasia) versus VVIQ \> 33 (typical imagers), following the conventions of
the time, and reported that aphantasics scored higher on two of the TAS-20's
three subscales. Likewise (even though they didn't compare VVIQ and TAS
explicitly), Ale & Burns used a VVIQ \< 32 threshold to defined aphantasia.
Redoing the comparison between groups on the combined Ale & Burns + Monzel et
al. sample, but this time splitting the aphantasia range itself — complete
aphantasia (VVIQ = 16) from hypophantasia (VVIQ 17-32, in the original,
provisional definition used at this stage) — told a different, more specific
story: it looked as though the *hypophantasia* sub-group was driving the
elevated scores, while complete aphantasics sat much closer to typical imagers
than the coarser two-group split had suggested.

Here is that first combined view, reproduced from the current, corrected data:

```{r ale-burns-monzel-plot, echo=FALSE}
early_data <- all_data |> dplyr::filter(study %in% c("burns", "monzel"))

library(patchwork)

plot_2g <- 
  plot_group_violins(
    tas ~ vviq_group_2,
    data = early_data,
    y_lab = "Total TAS score",
    base_size = 10
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(title = "The common 2-group comparison")

plot_4g <- 
  plot_group_violins(
    tas ~ vviq_group_4,
    data = early_data,
    y_lab = "Total TAS score",
    base_size = 10
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(title = "The finer-grained 4-group comparison")

plot_2g + plot_4g + 
  plot_layout(axis_titles = "collect") +
  plot_annotation(
    title = "Ale & Burns + Monzel et al. — the first combined look",
    subtitle = "The pattern that started everything",
    theme = theme(text = element_text(family ="Montserrat"))
  )
```

A negative relationship between visual imagery and alexithymia was already
visible from hypophantasia through hyperphantasia, but complete aphantasics (the
grey group) were "out of place": something different seemed to be happening
among people with a complete absence of imagery specifically.

## A finer split

Monzel et al.'s original paper had split their own sample at VVIQ \< 23
(aphantasia) versus VVIQ \> 33 (typical imagers), following the conventions of
the time, and reported that aphantasics scored higher on two of the TAS-20's
three subscales.

```{r early-refined-plot}
early_data_refined <- early_data |>
  dplyr::mutate(
    refined_group = dplyr::case_when(
      vviq == 16 ~ "Complete aphantasia",
      vviq <= 23 ~ "Hypophantasia",
      vviq >= 33 ~ "Typical imagers",
      TRUE ~ NA_character_
    )
  ) |>
  dplyr::filter(!is.na(refined_group))

plot_group_violins(
  tas ~ refined_group,
  data = early_data_refined,
  y_lab = "Total TAS score",
  base_size = 12
) +
  ggplot2::labs(
    title = "The finer split that changed the question",
    subtitle = "Complete aphantasia looks different from hypophantasia"
  )
```

This was the first sign that the eventual shape of the paper's finding —
complete absence of imagery behaving differently from merely weak imagery — was
not going to be a story that a single linear relationship, or even a simple
two-group split, could tell well.

## Ruby, analysed the same day

Perrine Ruby shared her dataset on 2025-10-29, and I ran the same analyses that
same day, using the package structure that had by then already taken shape from
the Ale & Burns / Monzel work. Ruby's dataset — 225 French participants (180
females, 42 males, 3 other; mean age 36, SD 16.1; range 10-82), collected as
part of a study on the sensory and emotional characteristics of autobiographical
and dream memories — also surfaced an incidental relationship between VVIQ and
dream recall frequency, which sits outside the scope of this study and was set
aside.

```{r ruby-plot}
data_with_ruby <- all_data |> dplyr::filter(study %in% c("burns", "monzel", "ruby"))

plot_group_violins(
  tas ~ vviq_group_4,
  data = data_with_ruby,
  y_lab = "Total TAS score",
  base_size = 12
) +
  scale_discrete_aphantasia() +
  ggplot2::labs(
    title = "Adding Ruby's data",
    subtitle = "The pattern holds with a third, independent sample"
  )
```

## An independent discovery, and a second correction

Midway through this process, Timo Kvamme and colleagues posted a preprint
reporting a related but more elaborate pattern in a large, independent sample:
opposing relationships between VVIQ and alexithymia either side of a fixed VVIQ
= 32 threshold — what became [Kvamme et al.
(2026)](https://doi.org/10.1016/j.neuropsychologia.2026.109368) once published.
I reached out, and Timo Kvamme generously agreed to share their data.

The same kind of data-quality check that had mattered for Ale & Burns' data
mattered again here, at a larger scale: I found that several TAS-20 items had
been reversed incorrectly in the shared dataset's preprocessing script. I
flagged this on 2025-11-12; Timo Kvamme confirmed the oversight and corrected it
the same day. Their final sample was substantial: 833 English-speaking
participants (426 females, 5 other genders; mean age 40.5, SD 13.4; range
18-83), recruited through Prolific and a pre-existing aphantasia database
specifically to cover the full range of imagery vividness.

```{r kvamme-plot}
data_with_kvamme <- all_data |> dplyr::filter(study %in% c("burns", "monzel", "ruby", "kvamme"))

plot_group_violins(
  tas ~ vviq_group_4,
  data = data_with_kvamme,
  y_lab = "Total TAS score",
  base_size = 12
) +
  scale_discrete_aphantasia() +
  ggplot2::labs(
    title = "Adding Kvamme et al.'s data",
    subtitle = "A fourth, much larger, independent sample — the pattern strengthens"
  )
```

Kvamme et al.'s own published analysis (a fixed split at VVIQ = 32, described in
full on [the model-comparison page](https://TODO-LINK-TO-KVAMME-PAGE)) reached a
related but distinct conclusion using a threshold chosen by hand, rather than
estimated from the data. That comparison — and what our own, data-driven
estimate of where the relationship actually changes shape turned out to say — is
one of the more interesting methodological threads in this whole project, and
gets its own full treatment later in this report.

## Mas & Luminet, completing the pool

The last dataset to arrive, on 2025-12-12, came from Marine Mas and Olivier
Luminet, collected as part of a preregistered lab study on alexithymia and
mental representations. 123 French-speaking participants (110 females; mean age
19.78, SD 1.15; range 18-24) were recruited from a research methods course.

```{r full-pool-plot}
plot_group_violins(
  tas ~ vviq_group_4,
  data = all_data,
  y_lab = "Total TAS score",
  base_size = 12
) +
  scale_discrete_aphantasia() +
  ggplot2::labs(
    title = "The complete pooled sample",
    subtitle = "Five studies, one consistent pattern"
  )
```

## From groundwork to a paper in its own right

None of this was originally meant to be the study. It was meant to be groundwork
— a way of estimating expected effect sizes before designing the larger
VVIQ/TAS/PsiQ/DERS/ERQ/PAQ/RRS study mentioned at the start of this page. It was
in discussing the pooled, cross-dataset pattern with Perrine Ruby that the
project changed shape: what had been preparation looked, on its own, like a
finding worth reporting — with the eventual four-group classification, and later
the non-linear modelling approach described on the [model comparison
page](https://TODO-LINK), as its distinguishing methodological contributions.

## Dead ends, for the record

Not every dataset search leads anywhere, and it is worth recording the ones that
did not — both because they are a real part of how this project came together,
and because they remain useful references for anyone wondering whether earlier
work exists on this specific combination of measures.

I contacted the authors of three "pre-aphantasia-era" (pre-2015) papers that had
used both the VVIQ and TAS-20, before the term "aphantasia" was in wide use:

- Wang & Yang (2024), *Mental Imagery in the Relationship between Alexithymia
  and Parental Psychological Control*
  ([doi.org/10.3390/bs14030183](https://doi.org/10.3390/bs14030183)) — no
  response.
- Jungmann et al. (2022), *Erfassung der Lebendigkeit mentaler
  Vorstellungsbilder*
  ([doi.org/10.1026/0012-1924/a000291](https://doi.org/10.1026/0012-1924/a000291))
  — no response.
- Leving (2024), an unpublished Master's thesis supervised by Jeanne Watson at
  the University of Toronto
  ([utoronto.scholaris.ca](https://utoronto.scholaris.ca/server/api/core/bitstreams/ea4d3c6b-34a2-4ffb-9ff1-e5be02218047/content))
  — I wrote to Professor Watson directly; no response.

I also contacted Alfredo Campos regarding Campos & Pérez (2000)'s *Alexithymia
and mental imagery*
([doi.org/10.1016/S0191-8869(99)00231-7](https://doi.org/10.1016/S0191-8869(99)00231-7)).
Professor Campos responded, but explained that he had lost contact with his
co-authors and had never held the data himself — so, more than two decades on,
that dataset is simply gone. It is a small, concrete reminder of what
data-sharing norms looked like before the open science practices this project
relies on became standard, and a large part of why this project's own data and
code are archived as thoroughly as they are (see the [OSF
project](https://doi.org/10.17605/OSF.IO/B837S) and this report's own [GitHub
repository](https://github.com/m-delem/aphantasiaEmotions)).
