---
title: "Codebook"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
```

This page documents every column in `all_data`, the pooled dataset described on
the [sample
description](https://m-delem.github.io/aphantasiaEmotions/articles/sample-description.html)
page. It is meant to serve the same purpose as the data dictionaries shared
alongside the flat `merged_data.xlsx` file archived on
[OSF](https://osf.io/), for anyone who wants to reuse the data without also
pulling in this package.

The tables below are computed directly from `all_data` rather than typed out
by hand, so they cannot drift out of sync with the data itself. A
machine-readable version of this same information, following the
[Psych-DS](https://psych-ds.github.io/) `dataset_description.json`
convention, is archived alongside the data on OSF.

## Core variables

```{r core-variables}
core_variables <- tibble::tribble(
  ~Variable,        ~Description,
  "study",          "Study of origin.",
  "lang",           "Language in which the study was administered.",
  "id",             "Unique participant identifier (\"subj_<study>_<number>\"), unique across the whole pooled dataset.",
  "sex",            "Sex assigned at birth, as reported by the participant.",
  "gender",         "Gender identity, as reported by the participant, where collected separately from sex.",
  "age",            "Participant age in years, self-reported.",
  "vviq",           "Total score on the VVIQ [@marksVisualImageryDifferences1973], summed across all 16 items. Lower scores indicate weaker/absent visual imagery.",
  "tas",            "Total score on the TAS-20 [@bagbyTwentyitemTorontoAlexithymia1994], summed across all 20 items (already reverse-scored where applicable). Higher scores indicate greater alexithymia.",
  "tas_identify",   "Difficulty Identifying Feelings (DIF) subscale of the TAS-20. Sum of items 1, 3, 6, 7, 9, 13, 14.",
  "tas_describe",   "Difficulty Describing Feelings (DDF) subscale of the TAS-20. Sum of items 2, 4, 11, 12, 17.",
  "tas_external",   "Externally Oriented Thinking (EOT) subscale of the TAS-20. Sum of items 5, 8, 10, 15, 16, 18, 19, 20.",
  "tas_group",      "Alexithymia status from the standard TAS-20 cutoff (total >= 61 = alexithymia).",
  "vviq_group_4",   "VVIQ group, 4 levels: aphantasia, hypophantasia, typical, hyperphantasia.",
  "vviq_group_3",   "VVIQ group, 3 levels: aphantasia, hypophantasia, typical (typical and hyperphantasia collapsed).",
  "vviq_group_2",   "VVIQ group, 2 levels: aphantasia, typical (broadest split)."
)

summarise_column <- function(data, colname) {
  x <- data[[colname]]
  n_missing <- sum(is.na(x))

  if (is.numeric(x)) {
    range_txt <- paste0(min(x, na.rm = TRUE), "-", max(x, na.rm = TRUE))
    type_txt <- "integer"
  } else {
    lvls <- if (colname!="id") sort(unique(x[!is.na(x)])) else "subj_[study]_[number]"
    range_txt <- paste(lvls, collapse = ", ")
    type_txt <- "string"
  }

  tibble::tibble(
    Variable = colname,
    Type = type_txt,
    `Range / levels` = range_txt,
    `N missing` = n_missing
  )
}

core_summary <- purrr::map_dfr(core_variables$Variable, \(v) summarise_column(all_data, v))

core_variables |>
  dplyr::left_join(core_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

## VVIQ item-level responses (`vviq_q1`-`vviq_q16`)

The VVIQ [@marksVisualImageryDifferences1973] has no subscales; its 16 items
are grouped into 4 scenes of 4 items each (a relative or friend you often see;
the rising sun; the front of a familiar shop; a country scene with trees,
mountains, and a lake), each rated 1 ("perfectly clear and as vivid as normal
vision") to 5 ("no image at all"). **Item wording is not reproduced here**, as
it is copyrighted by the scale's author — see @marksVisualImageryDifferences1973
for the original items.

```{r vviq-items}
vviq_scenes <- c(
  "a relative or friend you often see",
  "the rising sun",
  "the front of a familiar shop",
  "a country scene with trees, mountains, and a lake"
)

vviq_item_table <- tibble::tibble(item = 1:16) |>
  dplyr::mutate(
    Variable = paste0("vviq_q", item),
    Scene = vviq_scenes[ceiling(item / 4)]
  )

vviq_summary <- purrr::map_dfr(
  vviq_item_table$Variable,
  \(v) summarise_column(all_data |> tidyr::unnest(items, keep_empty = TRUE), v)
)

vviq_item_table |>
  dplyr::left_join(vviq_summary, by = "Variable") |>
  dplyr::select(Variable, Scene, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

Item-level VVIQ data are available for the burns, mas, ruby, and kvamme
studies. They are missing (`NA`) for all `r sum(all_data$study == "monzel")`
Monzel participants, for whom only total scores were shared.


## TAS-20 item-level responses (`tas_q1`-`tas_q20`)

The TAS-20 [@bagbyTwentyitemTorontoAlexithymia1994] has 20 items rated 1
("strongly disagree") to 5 ("strongly agree"), assigned to the three
sub-scales described above. **Item wording is not reproduced here**, as it is
copyrighted by the scale's authors — see @bagbyTwentyitemTorontoAlexithymia1994
for the original items.

Five items (4, 5, 10, 18, 19) are reverse-keyed on the original scale.
**Values in `tas_q4`, `tas_q5`, `tas_q10`, `tas_q18`, and `tas_q19` in
`all_data`/`merged_data.xlsx` are already reverse-scored** — summing
`tas_q1`-`tas_q20` directly reproduces the `tas` total column, with no
further reversal needed.

```{r tas-items}
tas_subscale_of <- function(i) {
  dplyr::case_when(
    i %in% c(1, 3, 6, 7, 9, 13, 14)      ~ "DIF",
    i %in% c(2, 4, 11, 12, 17)           ~ "DDF",
    i %in% c(5, 8, 10, 15, 16, 18, 19, 20) ~ "EOT"
  )
}

tas_reverse_scored <- c(4, 5, 10, 18, 19)

tas_item_table <- tibble::tibble(item = 1:20) |>
  dplyr::mutate(
    Variable = paste0("tas_q", item),
    Subscale = tas_subscale_of(item),
    `Reverse-keyed` = ifelse(item %in% tas_reverse_scored, "yes (already reversed)", "")
  )

tas_summary <- purrr::map_dfr(
  tas_item_table$Variable,
  \(v) summarise_column(all_data |> tidyr::unnest(items, keep_empty = TRUE), v)
)

tas_item_table |>
  dplyr::left_join(tas_summary, by = "Variable") |>
  dplyr::select(
    Variable, Subscale, `Reverse-keyed`, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

Item-level TAS-20 data are available for the same four studies as the VVIQ
items above, and missing for the same `r sum(all_data$study == "monzel")`
Monzel participants.

## A note on `sex` and `gender`

For most participants, `sex` and `gender` coincide. The Burns study collected
both constructs separately, and a small number of participants
(n = `r sum(all_data$sex != all_data$gender, na.rm = TRUE)`) reported a sex
and gender that differ.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page is a technical
reference, linked to from the narrative pages rather than meant to be read
start to finish. See the [sample
description](https://m-delem.github.io/aphantasiaEmotions/articles/sample-description.html)
page for how the five studies were combined, or [implementation
notes](https://m-delem.github.io/aphantasiaEmotions/articles/implementation-notes.html)
for the modelling code's conventions.

## References
