---
title: "The floor-group model, in depth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The floor-group model, in depth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
```

## Coming soon

This page will present the study's central finding in full: the floor-group
model's figure and coefficients, the pd/ROPE evidence behind it, why a
Gaussian family was used and checked, and how the finding holds up once
study-level heterogeneity is accounted for (the multilevel model, the
per-study breakdown, and the prior sensitivity check).

---

**Continuing through the Extended Online Report:** this page follows the
[model comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html).
To keep reading in order, continue to
[for those who come after](https://m-delem.github.io/aphantasiaEmotions/articles/for-those-who-come-after.html)
next.
