---
title: "The floor-group model, in depth"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{The floor-group model, in depth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
library(patchwork)

# Models and results are loaded directly from their saved artefacts in the 
# vignette, and explicitly never refitted (this is merely to protect the
# website). The "pkg" shorthand will be used throughout to point to the files of
# the aphantasiaEmotions package.
# See the Implementation Notes page for how these models were actually built.
pkg <- "aphantasiaEmotions"
refit <- "never"
```

The [previous
page](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
ended on a twist: among six candidate models, the one that best combines quality
of fit and parsimony is not the most flexible one (the segmented model with an
estimated knot), but the simplest one that captures the right structure: a
plain linear relationship among everyone above VVIQ = 16, plus a single
coefficient letting complete aphantasics have their own mean. This page is the
full case for that model: what it says, how confident we can be in what it says,
and whether it holds up once study-level heterogeneity and prior choice are both
accounted for.

## The model

```{r model-formula}
# Creating a binary column for whether a participant is in the floor-VVIQ group
# (complete aphantasia) or not
model_data <- all_data
model_data$complete_aphant <- factor(
  ifelse(model_data$vviq_group_4 == "aphantasia", "floor", "above_floor"),
  levels = c("above_floor", "floor")
)

# Fitting the model
floor_group_additive_multilevel <- fit_brms_model(
  formula = tas ~ vviq + complete_aphant + (vviq | study),
  data    = model_data,
  prior   = brms::prior(normal(0, 20), class = "b"),
  file    = system.file(
    "models", "floor_group_additive_multilevel_tot.rds", package = pkg),
  file_refit = refit
)
```

The formula is deliberately asymmetric, and that asymmetry is the whole point.
Complete aphantasics (VVIQ = 16) have no variance in VVIQ among themselves (every one of them has the same score) so there is no data to estimate a
VVIQ-TAS *slope* specific to that group. What the data can support is a single,
well-identified quantity: how far that group's mean TAS score sits from where
the continuous relationship, fit on everyone else, would have predicted it.
`complete_aphant` is that quantity. `(vviq | study)` lets both the slope and the
intercept vary by study, which is what makes the model's conclusions checkable
against between-study heterogeneity rather than just the pooled average (see
[below](#multilevel)).

## The floor group, visualised

Before the full figure, it's worth seeing the piece of data that motivates this
whole model on its own: the VVIQ distribution itself is not smoothly continuous.
There is a sharp, isolated spike at the scale's floor, distinct from a more
continuous, if irregular, remainder above it.

```{r vviq-marginal-standalone}
#| fig.alt: >
#|   A histogram of VVIQ scores across the full pooled sample, showing a
#|   sharp, isolated spike at the floor value (VVIQ = 16, coloured red)
#|   and a more continuous, irregular distribution of scores from about
#|   20 to 80 (coloured with a viridis gradient).

plot_vviq_marginal_histogram(model_data, base_size = 16) +
  ggplot2::labs(
    title = "The VVIQ distribution isn't smoothly continuous",
    x = "VVIQ score"
  )
```

That spike is the reason a group-specific intercept, rather than a smooth curve,
turns out to be the right tool: the data itself is telling you one part of the
range behaves like a distinct category, not like the tail of a continuum.

Composed with the model's own fitted relationship, the same histogram becomes
the header panel of this project's signature figure:

```{r floor-group-figure, fig.height=8}
#| fig.alt: >
#|   A two-panel composite figure. The top panel repeats the VVIQ
#|   histogram above. The bottom panel shows Total TAS score against
#|   VVIQ score, with a fitted line for the above-floor relationship
#|   (coloured by a viridis gradient matching the histogram), a violin
#|   plot showing the floor group's own TAS distribution to the left, a
#|   cross marking where the above-floor line would predict the floor
#|   group's mean to be, and an arrow showing the gap between that
#|   prediction and the floor group's actual mean.

hist_panel <-
  plot_vviq_marginal_histogram(model_data, base_size = 16) +
  ggplot2::geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  ggplot2::scale_x_continuous(
    limits = c(8, 81),
    expand = ggplot2::expansion(c(0.02, 0))
  )

main_panel <- 
  plot_floor_group(
    floor_group_additive_multilevel, model_data, 
    base_size = 16, stat_txt_size = 4.5)

hist_panel / main_panel + patchwork::plot_layout(heights = c(1, 4))
```

## The evidence

```{r floor-effect-evidence}
rope_range_contrast <- bayestestR::rope_range(floor_group_additive_multilevel)

sd_tas  <- stats::sd(model_data$tas)
sd_vviq <- stats::sd(model_data$vviq)
rope_range_slope <- 0.2 * (sd_tas / sd_vviq)

floor_effect <- bayestestR::describe_posterior(
  floor_group_additive_multilevel,
  parameters = "complete_aphant",
  rope_range = rope_range_contrast
)

vviq_slope <- bayestestR::describe_posterior(
  floor_group_additive_multilevel,
  parameters = "vviq",
  rope_range = c(-rope_range_slope, rope_range_slope)
)
```

The floor group's mean sits `r sprintf("%.2f", abs(floor_effect$Median))` points
below where the above-floor relationship would predict it (95% CI
[`r sprintf("%.2f", floor_effect$CI_low)`,
`r sprintf("%.2f", floor_effect$CI_high)`]), with a probability of direction of
`r sprintf("%.1f", floor_effect$pd * 100)`% and
`r sprintf("%.1f", floor_effect$ROPE_Percentage * 100)`% of the posterior inside
the region of practical equivalence to zero — as clear and as meaningful an
effect as this project's evidentiary standards can show.

The above-floor slope itself is `r sprintf("%.3f", vviq_slope$Median)` (95% CI
[`r sprintf("%.3f", vviq_slope$CI_low)`,
`r sprintf("%.3f", vviq_slope$CI_high)`]), pd =
`r sprintf("%.1f", vviq_slope$pd * 100)`%,
`r sprintf("%.1f", vviq_slope$ROPE_Percentage * 100)`% in ROPE. Note that this uses
a different ROPE convention than the floor-group contrast above, since a raw
slope and a group contrast aren't comparable on the same scale (see
[implementation
notes](https://m-delem.github.io/aphantasiaEmotions/articles/implementation-notes.html#rope-conventions-contrasts-vs-slopes)
for the full reasoning).

## Why gaussian()

Every model in this report, including this one, uses brms's default Gaussian
family. That choice is checked, not just assumed: see the [model
diagnostics](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html#gaussian-family)
page for the residual skewness, heteroscedasticity, and boundary checks behind
it.

## Multilevel robustness {#multilevel}

The result above already comes from the multilevel model: `(vviq | study)` is
part of the formula, not a separate add-on. It's worth showing directly what
that buys: does the floor effect look like a pooled-sample artefact, or does it
hold up study by study?

```{r per-study-figure}
#| fig.alt: >
#|   A line plot showing each of the five studies' own fitted VVIQ-TAS
#|   relationship (in five different colours, each restricted to that
#|   study's own observed VVIQ range), overlaid with the pooled
#|   population-level relationship in bold black. A muted violin at
#|   VVIQ=16 shows the floor group's pooled TAS distribution. Four of
#|   the five studies' lines cluster closely around the pooled line;
#|   one study (Mas & Luminet) diverges somewhat, with a shallower
#|   slope and a higher intercept.

study_coefs <- coef(floor_group_additive_multilevel)$study
study_coefs_df <- data.frame(
  study     = dimnames(study_coefs)[[1]],
  intercept = study_coefs[, "Estimate", "Intercept"],
  slope     = study_coefs[, "Estimate", "vviq"]
)

pooled_coefs <- brms::fixef(floor_group_additive_multilevel)

study_lines <- do.call(rbind, lapply(unique(model_data$study), function(s) {
  study_range <- range(model_data$vviq[model_data$study == s])
  coefs <- study_coefs_df[study_coefs_df$study == s, ]
  grid <- data.frame(
    vviq = seq(study_range[1], study_range[2], length.out = 100),
    study = s
  )
  grid$estimate <- coefs$intercept + coefs$slope * grid$vviq
  grid
}))

pooled_line <- data.frame(vviq = seq(16, 80, length.out = 100))
pooled_line$estimate <- pooled_coefs["Intercept", "Estimate"] +
  pooled_coefs["vviq", "Estimate"] * pooled_line$vviq

floor_raw <- model_data[model_data$vviq_group_4 == "aphantasia", ]
dens <- stats::density(floor_raw$tas, from = 20, to = 100, n = 200)
dens_scaled <- dens$y / max(dens$y) * 2
violin_df <- data.frame(x = 16 - dens_scaled, y = dens$x)

study_colors <- c(
  burns = "#E69F00", monzel = "#56B4E9", mas = "#009E73",
  ruby = "#D55E00", kvamme = "#CC79A7"
)
study_labels <-  c(
  burns = "Ale & B", monzel = "Monzel", mas = "Mas & L",
  ruby = "Ruby", kvamme = "Kvamme"
)

ggplot2::ggplot() +
  ggplot2::geom_point(
    data = model_data, ggplot2::aes(x = vviq, y = tas),
    alpha = 0.08, size = 0.8, color = "grey60"
  ) +
  ggplot2::geom_polygon(
    data = rbind(
      data.frame(x = violin_df$x, y = violin_df$y),
      data.frame(x = rep(16, nrow(violin_df)), y = rev(violin_df$y))
    ),
    ggplot2::aes(x = x, y = y),
    fill = "#C44E52", alpha = 0.35, color = "#8B3A3E", linewidth = 0.2
  ) +
  ggplot2::geom_line(
    data = pooled_line, ggplot2::aes(x = vviq, y = estimate),
    color = "black", linewidth = 1
  ) +
  ggplot2::geom_line(
    data = study_lines,
    ggplot2::aes(x = vviq, y = estimate, color = study),
    linewidth = 0.6, alpha = 0.85
  ) +
  ggplot2::scale_color_manual(
    values = study_colors, 
    labels = study_labels,
    name = "Study"
  ) +
  ggplot2::labs(
    x = "VVIQ score",
    y = "Total TAS score",
    caption = "Floor-group N by study: Ale & Burns: 60; Monzel et al: 20\nMas et Luminet: 0; Ruby: 13; Kvamme et al: 54"
  ) +
  theme_pdf(
    base_size = 16,
    base_theme = ggplot2::theme_minimal,
    panel.grid.minor = ggplot2::element_blank(),
    plot.caption = ggplot2::element_text(margin = ggplot2::margin(t = 10)),
    legend_relative = 0.9
  )
```

Four of the five studies' own fitted lines cluster closely around the pooled
relationship. Mas & Luminet's diverges somewhat — a shallower slope, offset
upward — consistent with that study's own composition described on the [sample
description](https://m-delem.github.io/aphantasiaEmotions/articles/sample-description.html)
page (young, homogeneous, no complete-aphantasia participants of its own to
anchor the floor group's contribution). The other four studies, spanning
different languages, recruitment channels, and sample sizes, tell a consistent
story.

### Prior sensitivity

The group-level slope SD term, i.e., how much the VVIQ-TAS slope is allowed to vary by study, relies on brms's own default weakly-informative prior rather than a
hand-picked one, deliberately: with only five studies informing that specific
variance component, a tighter, hand-chosen prior would risk doing more
inferential work than could be defended. The prior on fixed effects we chose was also deliberately weakly informative. Whether the model's substantive
conclusions depend on these choices is checked directly, refitting with priors
twice as wide as the defaults chosen:

```{r sensitivity-check}
sensitivity_priors <- c(
  brms::prior(
    normal(0, 40), class = "b"), # twice as wide as our normal(0,20) default
  brms::prior(
    student_t(3, 0, 26.6), # twice as wide as brms' default (13.3)
    class = "sd", group = "study", coef = "vviq")
)

floor_group_additive_multilevel_wide_prior <- fit_brms_model(
  formula = tas ~ vviq + complete_aphant + (vviq | study),
  data    = model_data,
  prior   = sensitivity_priors,
  file    = system.file(
    "models", "floor_group_additive_multilevel_wide_prior_tot.rds", 
    package = pkg),
  file_refit = refit
)

default_fixef <- brms::fixef(floor_group_additive_multilevel)
wide_fixef    <- brms::fixef(floor_group_additive_multilevel_wide_prior)

sensitivity_table <- data.frame(
  parameter = c("vviq (slope)", "complete_aphantfloor"),
  default_prior = c(
    default_fixef["vviq", "Estimate"], 
    default_fixef["complete_aphantfloor", "Estimate"]),
  wide_prior = c(
    wide_fixef["vviq", "Estimate"], 
    wide_fixef["complete_aphantfloor", "Estimate"])
)

sensitivity_table |> knitr::kable(digits = 3)
```

Both parameters are essentially unchanged between the default and the
deliberately wider priors: the headline result does not depend on which
weakly-informative priors were used to fit the model.

## TAS-20 sub-scales {#floor-subscales}

The total-TAS floor effect above is this project's central finding. The TAS-20
also has three established sub-scales — Difficulty Identifying Feelings (DIF),
Difficulty Describing Feelings (DDF), and Externally-Oriented Thinking (EOT) —
and the same model was fit separately on each, to check whether the floor effect
holds uniformly or is concentrated in a specific facet of alexithymia.

```{r subscale-results}
subscale_results <- readRDS(
  system.file("results", "floor_group_subscale_results.rds", package = pkg)
)

subscale_results |> knitr::kable(digits = 3)
```

```{r subscale-figures, fig.height=14}
#| fig.alt: >
#|   Three floor-group figures, one per TAS-20 subscale (Difficulty
#|   Identifying Feelings, Difficulty Describing Feelings, and
#|   Externally-Oriented Thinking), each following the same visual
#|   structure as the total-TAS figure above.

m_dif <- readRDS(
  system.file("models", "floor_group_additive_multilevel_dif.rds", package = pkg))
m_ddf <- readRDS(
  system.file("models", "floor_group_additive_multilevel_ddf.rds", package = pkg))
m_eot <- readRDS(
  system.file("models", "floor_group_additive_multilevel_eot.rds", package = pkg))

p_dif <- 
  plot_floor_group(
    m_dif, model_data, y_lab = "TAS DIF score",
    tas_breaks = scales::pretty_breaks(5),
    base_size = 16, 
    stat_txt_size = 4, 
    floor_label_size = 0,
    legend.position = "none") +
  ggplot2::labs(title = "Difficulty Identifying Feelings")

p_ddf <- 
  plot_floor_group(m_ddf, model_data, y_lab = "TAS DDF score",
    tas_breaks = scales::pretty_breaks(5),
    base_size = 16, 
    stat_txt_size = 4, 
    floor_label_size = 0,
    legend.position = "none") +
  ggplot2::labs(title = "Difficulty Describing Feelings")

p_eot <- 
  plot_floor_group(m_eot, model_data, y_lab = "TAS EOT score",
    tas_breaks = scales::pretty_breaks(5),
    base_size = 16, 
    stat_txt_size = 4, 
    floor_label_size = 0,
    legend.position = "none") +
  ggplot2::labs(title = "Externally-Oriented Thinking")

p_dif / p_ddf / p_eot
```

```{r subscale-interpretation-values}
#| include: false

get_val <- function(subscale, param, col) {
  subscale_results[
    subscale_results$subscale == subscale &
      subscale_results$parameter == param, col
  ]
}

dif_floor_rope <- get_val("DIF", "floor_effect", "pct_in_rope") * 100
ddf_floor_rope <- get_val("DDF", "floor_effect", "pct_in_rope") * 100
eot_floor_rope <- get_val("EOT", "floor_effect", "pct_in_rope") * 100

dif_slope_rope <- get_val("DIF", "vviq_slope", "pct_in_rope") * 100
ddf_slope_rope <- get_val("DDF", "vviq_slope", "pct_in_rope") * 100
eot_slope_rope <- get_val("EOT", "vviq_slope", "pct_in_rope") * 100
eot_slope_pd   <- get_val("EOT", "vviq_slope", "pd") * 100
```

The floor effect is unambiguous across all three sub-scales: the floor group's
mean sits clearly below the above-floor extrapolation on DIF, DDF, and EOT
alike, with `r sprintf("%.0f", dif_floor_rope)`%,
`r sprintf("%.0f", ddf_floor_rope)`%, and `r sprintf("%.0f", eot_floor_rope)`%
of each posterior distribution (respectively) inside its region of practical
equivalence to zero. This is not a pattern confined to one facet of alexithymia: complete aphantasics score in typical-imager territory across every sub-scale
this instrument distinguishes, not just on the total score.

The above-floor slopes tell a more textured story. DIF's and DDF's slopes are
both clearly outside their negligible-effect ranges
(`r sprintf("%.1f", dif_slope_rope)`% and `r sprintf("%.1f", ddf_slope_rope)`%
of their respective posteriors inside ROPE). EOT's slope is the one partial
exception: still directionally certain (pd = `r sprintf("%.1f", eot_slope_pd)`%)
and still mostly outside its ROPE, but with a meaningfully larger share of its
posterior (`r sprintf("%.1f", eot_slope_rope)`%) falling inside the negligible
range than either other sub-scale. In other words, the continuous relationship
between imagery vividness and alexithymia above the floor is more consistently
present for the difficulty-identifying and difficulty-describing facets than for
externally-oriented thinking specifically — while the floor effect itself, this
project's central finding, holds with equal force across all three.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page follows the [model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html).
To keep reading in order, continue to [for those who come
after](https://m-delem.github.io/aphantasiaEmotions/articles/for-those-who-come-after.html)
next. Or see [model
diagnostics](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html#ppc-floor-group)
and [implementation
notes](https://m-delem.github.io/aphantasiaEmotions/articles/implementation-notes.html)
for the technical detail behind this model.
