---
title: "For those who come after"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{For those who come after}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# NOTE: every figure-producing chunk needs a fig.alt option (pkgdown flags
# missing alt text; it is also good accessibility practice). Use Quarto-style
# #| pipe-comment chunk options throughout, as in how-this-study-found-its-shape.Rmd.

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
```

## Coming soon

This page will document the promising directions this project didn't have
time to pursue fully: a mixture-regression approach to the VVIQ floor spike,
a Bayesian MARS (BASS) exploration, and a per-study estimate of the segmented
model's breakpoint.

---

**This is the last page of the Extended Online Report.** If you've read
through from the start, thank you for following the whole story. You can
always jump back to
[how this study found its shape](https://m-delem.github.io/aphantasiaEmotions/articles/how-this-study-found-its-shape.html),
or return to the
[package homepage](https://m-delem.github.io/aphantasiaEmotions/).
