---
title: "How this study found its shape"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{How this study found its shape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

# NOTE FOR FUTURE VIGNETTES (sample-description, model-comparison,
# floor-group-model, for-those-who-come-after): every figure-producing chunk
# needs a fig.alt option - pkgdown flags figures missing alt text, and it's
# also just good accessibility practice for a public-facing report. Use the
# Quarto-style #| pipe-comment syntax for chunk options throughout, as done
# in this file, rather than inline ```{r, opt = val} syntax.

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
library(patchwork)
```

This study did not begin as a study. It began as a side question attached to a
different one, and the shape it ended up taking was not planned in advance so
much as discovered, one dataset at a time. Most papers describe what was found,
not how the finding was arrived at. This page exists to tell that second story:
the searching, the emailing, the re-analysing after every new dataset arrived.
This page is about the discovery process.

*A note on voice: this page is written in the first person throughout. That's
not a stylistic default: most of the searching, emailing, and re-analysing
described here happened as solo, iterative work between datasets, often before
there was a clear "we" to speak for. The manuscript itself credits contributions
properly in the conventional third person; this page simply reflects how the
work actually unfolded, day to day.*

## Where it started

I was preparing for a larger planned study on aphantasia and alexithymia,
together with Gaën Plancher, Marine Mas, and Olivier Luminet. That study was
going to use a fuller battery of questionnaires than the usual VVIQ (visual
imagery) and TAS-20 (alexithymia) alone, since we were, and remain, fairly
critical of what those two standard instruments actually capture on their own.

Before designing that study, I wanted to ground it in what was already known. So
the first, modest step was simply to see how much existing data on the VVIQ and
TAS-20 already existed, to get a sense of expected effect sizes and to plan
sensible sequential analyses.

## Ale and Burns, and a locked door

On October 13, 2025, I wrote to Edwin Burns about a preprint he and his
co-author had posted the year before, on aphantasia, alexithymia, and PTSD
symptomatology:
[doi.org/10.31234/osf.io/kj5d3](https://doi.org/10.31234/osf.io/kj5d3). Their
paper reported analyses of alexithymia and PTSD separately in relation to
aphantasia, but not the direct relationship between aphantasia and alexithymia
themselves, i.e., the co-occurrence question. I asked whether that had simply
not been reported, and mentioned that I'd been curious to look at their data
myself, except the OSF link in the paper pointed to a private repository.

Edwin Burns kindly shared the dataset. A short back-and-forth followed in which
I flagged what turned out to be a real discrepancy in how the TAS-20 total score
had been computed in the file — the kind of routine data-quality check that's
worth doing on any second-hand dataset before trusting it, and a good habit to
have gotten into early, since it would matter again later with a much larger
dataset. Edwin Burns confirmed the issue, and by October 29, the correction was
in place and the data were usable.

Ale and Burns' [-@aleAphantasiaAlexithymiaPredict2024] final sample: 192
English-speaking participants (122 females, 3 other genders; mean age 38.7, SD
11.4; range 18-86), recruited via social media, now openly archived on the OSF
of the present study (<https://doi.org/10.17605/OSF.IO/B837S>).

## Monzel, and the first pattern with a finer split

I collected the @monzelAffectiveProcessingAphantasia2024 dataset in the
meantime: 105 English-speaking participants (74 females; mean age 27.9, SD 9.29;
range 18-59), recruited via the *Aphantasia Research Project Bonn*'s participant
database, also openly archived on their own OSF project
([osf.io/y9c8g](https://osf.io/y9c8g/?view_only=1e6bd8670a3f4eacb1cf0f600343205e)).

Monzel et al.'s original paper had split their own sample at VVIQ \< 23
(aphantasia) versus VVIQ \> 33 (typical imagers), following the conventions of
the time, and reported that aphantasics scored higher on two of the TAS-20's
three subscales. Likewise (even though they didn't compare VVIQ and TAS
explicitly), Ale & Burns used a VVIQ \< 32 threshold to defined aphantasia.
Redoing the comparison between groups on the combined Ale & Burns + Monzel et
al. sample, but this time splitting the aphantasia range itself — complete
aphantasia (VVIQ = 16) from hypophantasia (VVIQ 17-32) — told a different, more
specific story: it looked as though the *hypophantasia* sub-group was driving
the elevated scores, while complete aphantasics sat much closer to typical
imagers than the coarser two-group split had suggested.

Here is that first combined view, reproduced from the current, corrected data:

```{r ale-burns-monzel-plot}
#| fig.alt: >
#|   Two side-by-side raincloud plots comparing Total TAS score across VVIQ
#|   groups in the combined Ale & Burns and Monzel et al. sample. The left
#|   panel shows the common 2-group split (aphantasia vs. typical imagers);
#|   the right panel shows the finer-grained 4-group split (aphantasia,
#|   hypophantasia, typical, hyperphantasia). The 4-group panel reveals that
#|   complete aphantasics have lower alexithymia scores than hypophantasics,
#|   a distinction the 2-group split obscures.
early_data <- all_data |> dplyr::filter(study %in% c("burns", "monzel"))

plot_2g <- 
  plot_group_violins(
    tas ~ vviq_group_2,
    data = early_data,
    y_lab = "Total TAS score",
    base_size = 16
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(title = "The common 2-group comparison")

plot_4g <- 
  plot_group_violins(
    tas ~ vviq_group_4,
    data = early_data,
    y_lab = "Total TAS score",
    base_size = 16
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(title = "The finer-grained 4-group comparison")

plot_2g + plot_4g + 
  plot_layout(axis_titles = "collect") +
  plot_annotation(
    title = "Ale & Burns + Monzel et al. — the first combined look (N = 297)",
    subtitle = "The pattern that started everything",
    theme = theme(text = element_text(family ="Montserrat", size = 20))
  )
```

A negative relationship between visual imagery and alexithymia was already
visible from hypophantasia through hyperphantasia, but complete aphantasics (the
right panel's grey group) were "out of place": something different seemed to be
happening among people with a complete absence of imagery specifically. The
common 2-group split (left panel) does not show this at all: it simply averages
complete aphantasics in with everyone below the threshold, hiding exactly the
distinction that turned out to matter.

This was the first sign that the eventual shape of the paper's finding —
complete absence of imagery behaving differently from merely weak imagery — was
not going to be a story that a single linear relationship, or even a simple
two-group split, could tell well.

## Ruby, analysed the same day

Perrine Ruby shared her dataset on October 29, 2025, and I ran the same analyses
that same day, using the package structure that had by then already taken shape
from the Ale & Burns / Monzel et al. work. Ruby's dataset — 225 French
participants (180 females, 42 males, 3 other; mean age 36, SD 16.1; range
10-82), collected as part of a study on the sensory and emotional
characteristics of autobiographical and dream memories — also surfaced an
incidental relationship between VVIQ and dream recall frequency, which sits
outside the scope of this study and was set aside.

```{r ruby-plot}
#| fig.alt: >
#|   A raincloud plot showing Total TAS score across the four VVIQ groups
#|   (aphantasia, hypophantasia, typical, hyperphantasia), now including
#|   Perrine Ruby's dataset alongside Ale & Burns and Monzel et al. The
#|   pattern seen in the first two datasets — complete aphantasics scoring
#|   lower than hypophantasics — holds with this third, independent sample.
data_with_ruby <- all_data |> dplyr::filter(study %in% c("burns", "monzel", "ruby"))

plot_group_violins(
  tas ~ vviq_group_4,
  data = data_with_ruby,
  y_lab = "Total TAS score",
  base_size = 16
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(
    title = "Adding Ruby's data (total N = 522)",
    subtitle = "The pattern holds with a third, independent sample"
  )
```

Even with Ruby's data, a large, independent sample with another native language
and sampling methods (no specific focus on aphantasia initially), the pattern
held.

## An independent discovery, and a second correction

Midway through this process, Timo Kvamme and colleagues posted a preprint
reporting a very similar pattern in a large, independent sample: opposing
relationships between VVIQ and alexithymia either side of a fixed VVIQ = 32
threshold, what became @kvammeWhenWeakImagery2026 once published. I reached out,
and Timo Kvamme generously agreed to share their data.

The same kind of data-quality check that had mattered for Ale & Burns' data
mattered again here, at a larger scale: on November 12, 2025, I noticed that
several TAS-20 items had been reversed incorrectly in the shared dataset's
preprocessing script. Timo Kvamme confirmed the oversight and corrected it the
same day. Their final sample was substantial: 833 English-speaking participants
(426 females, 5 other genders; mean age 40.5, SD 13.4; range 18-83), recruited
through Prolific and a pre-existing aphantasia database specifically to cover
the full range of imagery vividness.

```{r kvamme-plot}
#| fig.alt: >
#|   A raincloud plot showing Total TAS score across the four VVIQ groups,
#|   now including Kvamme et al.'s large independent sample alongside Ale &
#|   Burns, Monzel et al., and Ruby. Adding this fourth, much larger dataset
#|   strengthens the same pattern seen in the smaller samples.
data_with_kvamme <- all_data |> dplyr::filter(study %in% c("burns", "monzel", "ruby", "kvamme"))

plot_group_violins(
  tas ~ vviq_group_4,
  data = data_with_kvamme,
  y_lab = "Total TAS score",
  base_size = 16
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(
    title = "Adding Kvamme et al.'s data (new total N = 1355)",
    subtitle = "A fourth, much larger, independent sample — the pattern strengthens"
  )
```

Kvamme et al.'s own published analysis (a fixed split at VVIQ = 32, described in
full on [the model comparison
page](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html))
reached a related but distinct conclusion using a threshold chosen by hand,
rather than estimated from the data. That comparison, and what our own,
data-driven estimate of where the relationship actually changes shape turned out
to say, is one of the more interesting methodological threads in this whole
project, and gets its own full treatment later in this Extended Online Report.

## Mas & Luminet, completing the pool

The last dataset to arrive, on December 12, 2025, came from Marine Mas and
Olivier Luminet, collected as part of a preregistered lab study on alexithymia
and mental representations [@masPreregistration2025]. 123 French-speaking
participants (110 females; mean age 19.78, SD 1.15; range 18-24) were recruited
from a research methods course. The particularity of this dataset is that it was
collected without any aphantasia-targeted recruitment (Ruby's did some at the
end), and as a consequence contains only 2 hypophantasics, 1 hyperphantasic and
no complete aphantasics. It exhibited the negative VVIQ-TAS relationship on its
VVIQ range, and further strengthened the pattern on that part of the continuum.

```{r full-pool-plot}
#| fig.alt: >
#|   A raincloud plot showing Total TAS score across the four VVIQ groups
#|   in the complete pooled sample of 1478 participants from all five
#|   studies (Ale & Burns, Monzel et al., Ruby, Kvamme et al., and Mas &
#|   Luminet). The pattern is consistent across the whole pool: complete
#|   aphantasics score lower on alexithymia than hypophantasics.
plot_group_violins(
  tas ~ vviq_group_4,
  data = all_data,
  y_lab = "Total TAS score",
  base_size = 16
  ) +
  scale_x_aphantasia(add = c(0.4, 0.7)) +
  scale_discrete_aphantasia() +
  ggplot2::labs(
    title = "The complete pooled sample (N = 1478)",
    subtitle = "Five studies, one consistent pattern"
  )
```

## From groundwork to a paper in its own right

None of this was originally meant to be the study. It was meant to be
groundwork: a way of estimating expected effect sizes before designing the
larger study (with 7 different questionnaires!) mentioned at the start of this
page. It was in discussing the pooled, cross-dataset pattern with Perrine Ruby
that the project changed shape: what had been preparation looked, on its own,
like a finding worth reporting — with the eventual four-group classification,
and later the non-linear modelling approach described on the [model comparison
page](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html),
as its distinguishing methodological contributions.

## Dead ends, for the record

Not every dataset search leads anywhere, and it is worth recording the ones that
did not — both because they are a real part of how this project came together,
and because they remain useful references for anyone wondering whether earlier
work exists on this specific combination of measures.

I contacted the authors of three "pre-aphantasia-era" (pre-2015) papers that had
used both the VVIQ and TAS-20, before the term "aphantasia" was in wide use:

- Wang & Yang (2024), *Mental Imagery in the Relationship between Alexithymia
  and Parental Psychological Control*,
  [doi.org/10.3390/bs14030183](https://doi.org/10.3390/bs14030183). N = 282; the
  authors reported significant Pearson correlation of -.44 across the whole
  sample, -.28 in the alexithymic group alone, and -.21 in the non-alexithymic
  group. No distribution of VVIQ scores was reported, so no way of knowing
  whether or not extreme imagery groups were present. No response from the
  authors.
- Jungmann et al. (2022), *Erfassung der Lebendigkeit mentaler
  Vorstellungsbilder*,
  [doi.org/10.1026/0012-1924/a000291](https://doi.org/10.1026/0012-1924/a000291).
  N = 300; the authors reported (in their supplementary materials) weak but
  significant correlations of -.17, -.14 and -.15, -.13 between the TAS and the
  total VVIQ, "Person", "Shop" and "Landscape" groups of items respectively
  (relationship with the "Sunrise" group was a non-significant -.07). No
  distribution of VVIQ scores either. No response from the corresponding author.
- Leving (2024), an unpublished Master's thesis supervised by Jeanne Watson at
  the University of Toronto
  ([utoronto.scholaris.ca](https://utoronto.scholaris.ca/server/api/core/bitstreams/ea4d3c6b-34a2-4ffb-9ff1-e5be02218047/content)).
  N = 62; she reported a significant -.59 correlation between the TAS and VVIQ.
  No distribution of VVIQ scores either. I wrote to Professor Watson directly;
  no response.

I also contacted Alfredo Campos regarding Campos et al.'s (2000) *Alexithymia
and mental imagery*
([doi.org/10.1016/S0191-8869(99)00231-7](https://doi.org/10.1016/S0191-8869(99)00231-7)).
N = 133; the authors reported a significant -.23 correlation between the VVIQ
and total TAS scores, but found that dividing by TAS subscale showed that the
significant correlation held only for the *Difficulty describing feelings*
(-.23), but not with *Difficulty identifying feelings* (-.19) or *Externally
oriented thinking* (-.19)*.* Professor Campos responded, but explained that he
had lost contact with his co-authors and had never held the data himself: so,
more than two decades on, that dataset is simply gone. It is a small, concrete
reminder of what data-sharing norms looked like before the open science
practices this project relies on became standard, and a large part of why this
project's own data and code are archived as thoroughly as they are (see the [OSF
project](https://doi.org/10.17605/OSF.IO/B837S) and this report's own [GitHub
repository](https://github.com/m-delem/aphantasiaEmotions)).

## Contact me!

If you happen to know other studies that used these two questionnaires together,
feel free to [contact me](mailto:mael.delem@pm.me), I'd be happy to update the
analyses with new data.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this is the first page. To
keep reading in order, continue to the [sample
description](https://m-delem.github.io/aphantasiaEmotions/articles/sample-description.html)
next. Or jump straight to the [model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html),
[floor-group
model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html),
[model
diagnostics](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html),
[implementation
notes](https://m-delem.github.io/aphantasiaEmotions/articles/implementation-notes.html),
or [for those who come
after](https://m-delem.github.io/aphantasiaEmotions/articles/for-those-who-come-after.html).

--------------------------------------------------------------------------------

### References
