---
title: "Implementation notes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Implementation notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  warning = FALSE,
  message = FALSE
)
```

This page is for anyone who wants to build on this project's code directly:
replicate a model, adapt a formula, or understand a design choice that isn't
visible from the outside. None of what follows is necessary to trust the study's
conclusions (see the [model
diagnostics](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html)
page for that); it's here because getting some of these models to fit correctly
took real, sometimes non-obvious work, and that work is worth recording rather
than quietly smoothing over. This iterative work on getting the models right was
conducted in several (thoroughly commented) scripts in the
[`inst/scripts/`](https://github.com/m-delem/aphantasiaEmotions/tree/main/inst/scripts)
folder available in the package's source files. This page summarises some of the
core decisions taken in these scripts.

## Iteration and chain conventions

Every model in this project is fit with `fit_brms_model()`, a thin wrapper
around `brms::brm()`. Its `iterations` argument means **post-warmup draws per
chain**, not a total divided across chains. An earlier version of this function
divided a total iteration count by the number of detected CPU cores, which meant
the sampler's actual per-chain budget depended on whatever machine happened to
run it. That's not reproducible in any meaningful sense, so the function was
rewritten to take `chains`, `iterations`, and `warmup` as independent, explicit
arguments.

```{r fit-brms-model-signature}
fit_brms_model(
  ...,
  chains = 4,
  iterations = 2000,  # post-warmup draws PER CHAIN
  warmup = 1000,
  cores = chains,
  adapt_delta = 0.95,
  max_treedepth = 10,
  seed = 667
)
```

Most models in this project use `chains = 6` rather than the default 4, for
extra convergence confidence on the newer, less-standard model forms (the
segmented and floor-group models) at low extra cost on a machine with enough
cores to run them in parallel. This amounts to 12,000 post-warmup draws per
model to estimate its parameters.

## The segmented model's estimated knot

The estimated-knot segmented model
(`tas ~ a + b1 * vviq + b2 * (vviq - k) * step(vviq - k)`, fit as a nonlinear
brms model with `k` itself a free parameter) took two real rounds of debugging
before it produced trustworthy results.

**Problem 1: `fmin`/`fmax` compile in Stan but don't exist in R.** The first
version of this formula used `fmin(vviq, k)` and `fmax(vviq - k, 0)`, which are
valid Stan functions, and the model compiled and sampled without any error. The
problem only appeared afterwards: `brms::loo()` and `fitted()` need to
re-evaluate the non-linear formula in R for post-processing, and R has no
`fmin`/`fmax`. The fix was reformulating the hinge using `step()` instead, a
function that exists and behaves identically in both Stan and R:

```{r step-reformulation}
# Does NOT work post-fit (fmin/fmax don't exist in R):
# tas ~ a + b1 * fmin(vviq, k) + b2 * fmax(vviq - k, 0)

# Works in both contexts:
tas ~ a + b1 * vviq + b2 * (vviq - k) * step(vviq - k)
```

Under this re-parametrisation, `b1` is the below-knot slope directly, and
`b1 + b2` (not `b2` alone) is the above-knot slope — `b2` is the *change* in
slope at the knot, not a segment slope on its own. Worth remembering if you're
reading the raw coefficients rather than the derived slopes reported on the
[model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
page.

**Problem 2: bad default initialisation.** Even after switching to `step()`, a
small test fit found the sampler landing on `k = -3.39`, a knot location outside
VVIQ's actual range (16-80) entirely, with catastrophic effective sample size.
brms's default initialisation for non-linear parameters is naive (near zero,
with small random jitter), and zero is nowhere near a plausible knot location on
this scale. The fix was supplying explicit starting values, centred on sensible
guesses rather than left to chance:

```{r segmented-inits}
segmented_inits <- function() {
  list(
    b_a  = array(mean(all_data$tas)),
    b_b1 = array(0),
    b_b2 = array(0),
    b_k  = array(24)  # matches earth::earth()'s knot estimate
  )
}
```

With good starting values, the same model converged cleanly, and its estimated
knot (median 19.5, 95% CI [17.7, 24.1]) was stable whether fit with 200 draws on
one chain or the full 6-chain, 2000-draws-per-chain production run, which is a
useful cross-check that the fix addressed the real problem rather than just
moving it.

The model also needed `adapt_delta = 0.99` and `max_treedepth = 15` (both above
`fit_brms_model()`'s defaults) to fully clear divergent-transition warnings.
Non-linear/hinge models tend to have trickier posterior geometry than standard
linear ones, and this is a reasonable, low-cost precaution for that model class
specifically.

## ROPE conventions: contrasts vs. slopes

This project uses two different Region of Practical Equivalence (ROPE)
conventions, deliberately, not inconsistently.

For **group contrasts** (like the floor-group model's `complete_aphant`
coefficient), `bayestestR::rope_range()`'s default — 0.1 times the SD of the
outcome — is appropriate, since a group contrast is directly comparable to a
mean difference.

For **slopes** (the `vviq` coefficient in any model, or the segmented model's
below/above-knot slopes), that same convention is not appropriate: a raw slope
is expressed "per one unit of VVIQ," and VVIQ's scale (16-80) is arbitrary
relative to TAS's scale. Comparing a raw slope directly to `0.1 * SD(tas)`
implicitly assumes one unit of VVIQ is a standardised step, which it isn't.
Slopes in this project instead use a Cohen-motivated threshold: a standardised
effect of 0.2 is considered a small-to-noticeable effect, rescaled into the raw
units the slope is actually expressed in:

```{r rope-for-slopes}
sd_tas  <- sd(all_data$tas)
sd_vviq <- sd(all_data$vviq)

rope_range_slope <- 0.2 * (sd_tas / sd_vviq)
```

This distinction is worth getting right specifically because it's easy to miss:
both quantities look like they should use "the same" ROPE, and only looking
closely at what a slope versus a contrast actually represents makes clear why
they can't.

## The `earth`-derived fixed knot

The fixed-knot segmented model (a simpler, non-estimated companion to the model
discussed above) uses a knot location found by `earth::earth()`, extracted
programmatically rather than typed in by hand, so the code stays correct if the
underlying data ever changes:

```{r earth-knot-extraction}
mars <- earth::earth(tas ~ vviq, data = all_data)

# mars$cuts is a matrix of breakpoints used by each hinge term; for a
# single-knot model, all non-zero entries should agree.
vviq_cuts <- mars$cuts[, "vviq"]
knot <- unique(vviq_cuts[vviq_cuts != 0])
```

This fixed-knot model and the estimated-knot model above produce nearly
identical fits (LOO elpd difference of -0.09, well within noise), which is a
nice cross-validation of both approaches, using two genuinely different methods
to arrive at essentially the same answer.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page is a technical
reference, linked to from the narrative pages rather than meant to be read start
to finish. Return to the [model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
or [floor-group
model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html)
pages, or see [model
diagnostics](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html)
for convergence and predictive checks.
