---
title: "Model comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
```

## Coming soon

This page will walk through the full model comparison arc — from the naive
2-group threshold, through the categorical 4-group model, to the continuous
linear, GAM, and segmented models — including the direct comparison against
Kvamme et al.'s own published, fixed-threshold approach.

---

**Continuing through the Extended Online Report:** this page follows the
[sample description](https://m-delem.github.io/aphantasiaEmotions/articles/sample-description.html).
To keep reading in order, continue to
[the floor-group model, in depth](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html)
next. Or jump to
[for those who come after](https://m-delem.github.io/aphantasiaEmotions/articles/for-those-who-come-after.html).
