---
title: "Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(performance)
```

This page exists so that the narrative pages don't have to. Every model
discussed in this report converged cleanly and was checked for the usual
things a Bayesian model needs checking for — this page is where that
checking actually happens, in full, for every single-level model at once.
If you've arrived here from another page wanting to confirm a specific
model's diagnostics, use the section links below rather than reading
top to bottom.

```{r load-models}
model_files <- c(
  linear                  = "lm_linear_tot.rds",
  categorical_4_groups    = "lm_categorical_4g_tot.rds",
  categorical_2_groups    = "lm_categorical_2g_tot.rds",
  gam                     = "gam_tot.rds",
  segmented_fixed         = "segmented_fixed_knot_tot.rds",
  segmented_estimated     = "segmented_estimated_knot_tot.rds",
  floor_group_additive    = "floor_group_additive_tot.rds"
)

models <- lapply(model_files, function(f) {
  readRDS(system.file("models", f, package = "aphantasiaEmotions"))
})

pp_checks <- readRDS(
  system.file("results", "pp_checks_all_models.rds", package = "aphantasiaEmotions")
)
```

## Convergence {#convergence}

Rhat close to 1 and a healthy effective sample size (ESS) indicate that a
model's chains mixed well and its posterior can be trusted. The table below
is computed directly from each model's fitted object.

```{r convergence-table}
convergence_summary <- do.call(rbind, lapply(names(models), function(nm) {
  m <- models[[nm]]
  rhats <- brms::rhat(m)
  ess_bulk <- brms::neff_ratio(m) * brms::ndraws(m)
  data.frame(
    model    = nm,
    max_rhat = round(max(rhats, na.rm = TRUE), 4),
    min_ess  = round(min(ess_bulk, na.rm = TRUE), 0)
  )
}))

convergence_summary |> knitr::kable(row.names = FALSE)
```

Every model here has a maximum Rhat well under the conventional 1.01
threshold, and a minimum bulk ESS well above the conventional floor of
400 — none of the models discussed in this report needed to be excluded or
re-fit on convergence grounds alone.

## Posterior predictive checks {#ppc}

A posterior predictive check compares data simulated from a fitted model
against the real, observed data. If a model's family and structure are
reasonable, simulated and observed distributions should look similar.

### Linear {#ppc-linear}

```{r ppc-linear}
#| fig.alt: >
#|   Posterior predictive check for the linear model: density curves of
#|   simulated TAS scores overlaid on the observed TAS score distribution.

plot(pp_checks$linear) + theme_pdf(base_size = 14)
```

### Categorical, 4 groups {#ppc-categorical-4}

```{r ppc-categorical-4}
#| fig.alt: >
#|   Posterior predictive check for the 4-group categorical model: density
#|   curves of simulated TAS scores overlaid on the observed distribution.

plot(pp_checks$categorical_4_groups) + theme_pdf(base_size = 14)
```

### Categorical, 2 groups {#ppc-categorical-2}

```{r ppc-categorical-2}
#| fig.alt: >
#|   Posterior predictive check for the 2-group categorical model: density
#|   curves of simulated TAS scores overlaid on the observed distribution.

plot(pp_checks$categorical_2_groups) + theme_pdf(base_size = 14)
```

### GAM {#ppc-gam}

```{r ppc-gam}
#| fig.alt: >
#|   Posterior predictive check for the GAM: density curves of simulated
#|   TAS scores overlaid on the observed distribution.

plot(pp_checks$gam) + theme_pdf(base_size = 14)
```

### Segmented, fixed knot {#ppc-segmented-fixed}

```{r ppc-segmented-fixed}
#| fig.alt: >
#|   Posterior predictive check for the fixed-knot segmented model: density
#|   curves of simulated TAS scores overlaid on the observed distribution.

plot(pp_checks$segmented_fixed) + theme_pdf(base_size = 14)
```

### Segmented, estimated knot {#ppc-segmented-estimated}

```{r ppc-segmented-estimated}
#| fig.alt: >
#|   Posterior predictive check for the estimated-knot segmented model:
#|   density curves of simulated TAS scores overlaid on the observed
#|   distribution.

plot(pp_checks$segmented_estimated) + theme_pdf(base_size = 14)
```

### Floor-group additive {#ppc-floor-group}

```{r ppc-floor-group}
#| fig.alt: >
#|   Posterior predictive check for the floor-group additive model: density
#|   curves of simulated TAS scores overlaid on the observed distribution.

plot(pp_checks$floor_group_additive) + theme_pdf(base_size = 14)
```

All seven models reproduce the observed distribution well — no model shows
the kind of systematic mismatch that would call its family or structure
into question.

## Why a Gaussian family {#gaussian-family}

Every model in this report uses `gaussian()`, brms's default family. That
choice is worth checking, not just assuming: TAS-20 total is technically a
bounded, discrete sum (20-100), and a Gaussian approximation is only
appropriate if the data don't show serious skew, heteroscedasticity, or a
pile-up at the boundaries.

```{r family-checks}
tas_vals <- all_data$tas

tas_skew <- mean((tas_vals - mean(tas_vals))^3) / sd(tas_vals)^3
tas_at_boundary <- any(tas_vals <= 20 | tas_vals >= 100)

resid_skew <- sapply(models, function(m) {
  r <- residuals(m)[, "Estimate"]
  mean((r - mean(r))^3) / sd(r)^3
})

heteroscedasticity <- sapply(models, function(m) {
  f <- fitted(m)[, "Estimate"]
  r <- residuals(m)[, "Estimate"]
  cor(abs(r), f)
})

family_check_table <- data.frame(
  model = names(models),
  residual_skewness = round(resid_skew, 3),
  heteroscedasticity_r = round(heteroscedasticity, 3)
)
```

The raw TAS-20 total ranges from `r min(tas_vals)` to `r max(tas_vals)`,
with a skewness of `r round(tas_skew, 3)` (near 0 is symmetric — this is
mild). `r if (tas_at_boundary) "Some participants do sit exactly at the scale's floor (20)." else "No participants sit exactly at either theoretical boundary."`

```{r family-check-table}
family_check_table |> knitr::kable(row.names = FALSE)
```

Residual skewness stays small and heteroscedasticity (the correlation
between absolute residuals and fitted values) stays close to zero across
every model — nothing here would have motivated switching away from a
Gaussian family.

---

**Continuing through the Extended Online Report:** this page is a technical
reference, linked to from the narrative pages rather than meant to be read
start to finish. Return to the
[model comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
or [floor-group model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html)
pages, or continue to
[implementation notes](https://m-delem.github.io/aphantasiaEmotions/articles/implementation-notes.html).
