---
title: "Sample description"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Sample description}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
library(ggpubr)
```

Every analysis in this report runs on `all_data`, containing `r nrow(all_data)`
participants pooled from five studies: two unpublished datasets shared by the
authors specifically for this project, three retrieved from previously published
or shared work. This page describes where each of the five came from, who was in
them, and a few places where they genuinely differ from each other. Differences
that may matter later, when the question becomes whether a pattern found in the
pooled sample holds up once study membership is accounted for (see the
[floor-group
model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.htm#multilevel)
page).

## The five studies

```{r sample-table}
all_data |>
  dplyr::group_by(study) |>
  dplyr::reframe(
    Language = unique(lang),
    N = paste0(
      dplyr::n(),
      " (",
      sum(gender == "female", na.rm = TRUE),
      " F, ",
      sum(gender == "other", na.rm = TRUE),
      " O)"
    ),
    M_age = mean(age, na.rm = TRUE),
    SD_age = sd(age, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE)
  ) |>
  knitr::kable(digits = 2)
```

**Ale & Burns** [-@aleAphantasiaAlexithymiaPredict2024] collected VVIQ and
TAS-20 data as part of a study on aphantasia, alexithymia, and PTSD
symptomatology. 192 English-speaking participants (122 females, 3 other genders;
mean age 38.7, SD 11.4; range 18-86) were recruited via social media. Their data
are archived on a private OSF ([osf.io/hqz3e](https://osf.io/hqz3e/)), but made
openly available in clean form in this [study's own OSF
project](https://doi.org/10.17605/OSF.IO/B837S).

**Monzel et al.** [-@monzelAffectiveProcessingAphantasia2024] collected VVIQ and
TAS-20 data as part of a study on aphantasia, alexithymia, and affective
processing. 105 English-speaking participants (74 females; mean age 27.9, SD
9.29; range 18-59) were recruited via the *Aphantasia Research Project Bonn*'s
participant database, split into 75 controls (VVIQ \> 32) and 30 participants
with weak or no imagery (VVIQ ≤ 23). Their own paper reports a real,
worth-flagging imbalance: the aphantasia sub-group was on average 6.6 years
older than controls (t = 3.09, p = .004, d = 0.76), a difference large enough that the
original authors treated age as a covariate throughout. It's a genuine
limitation of that specific sample, though not one that appears to have
distorted the pattern this project is built on: the VVIQ-TAS relationship in
Monzel et al.'s data looks the same as in the other four studies. Their data are
openly archived on OSF
([osf.io/y9c8g](https://osf.io/y9c8g/?view_only=1e6bd8670a3f4eacb1cf0f600343205e)).

**Ruby** collected VVIQ and TAS-20 data as part of a study on the sensory and
emotional characteristics of autobiographical and dream memories. 225 French
participants (180 females, 42 males, 3 other; mean age 36, SD 16.1; range
10-82), recruited by an announcement on social media and on mailing lists
dedicated to research volunteers in Lyon and Paris, completed VVIQ and TAS-20
together with sensory and emotional scales to describe the one autobiographical
memory from the day before and the one dream memory from the night before they
had to report. The inclusion criterion was to have a memory of a dream from the
night before when filling in the questionnaire.

**Mas & Luminet** [-@masPreregistration2025] collected VVIQ and TAS-20 data as
part of a preregistered lab experiment on alexithymia and mental
representations. 123 French-speaking participants (110 females; mean age 19.78,
SD 1.15; range 18-24) were recruited from a research methods course, in exchange
for course credit. No inclusion or exclusion criteria were specified for this
study.

**Kvamme et al.** [-@kvammeWhenWeakImagery2026] collected VVIQ and TAS-20 data
as part of a much larger study on mental imagery, mental health, subjective
interoception, and alexithymia. 833 English-speaking participants (426 females,
5 other genders; mean age 40.5, SD 13.4; range 18-83) were recruited through
Prolific, in two phases: an initial, targeted recruitment of individuals with
VVIQ ≤ 32 from a pre-existing database, followed by open recruitment to cover
the full range of imagery vividness. That two-phase design is a large part of
why this study contributes such a substantial share of the pool's
complete-aphantasia and hypophantasia participants despite not being an
aphantasia-only sample (see the [model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
page for how much that group specifically matters to this project's central
finding).

Kvamme et al.'s exclusion pipeline is also worth naming directly, since it is
more thoroughly documented than most: of an initial 855 respondents, 22 were
excluded for excessively short or implausibly long completion times, and for
suspected careless responding, detected with the Even-Odd Inconsistency Index
(via the `careless` R package). The 833 remaining participants are what this
project's `kvamme` study component contains.

## By VVIQ group

The same pooled sample, seen instead through the four-group VVIQ classification
used throughout this report:

```{r vviq-group-table}
all_data |>
  dplyr::group_by(vviq_group_4) |>
  dplyr::reframe(
    N = paste0(
      dplyr::n(),
      " (",
      sum(gender == "female", na.rm = TRUE),
      " F, ",
      sum(gender == "other", na.rm = TRUE),
      " O)"
    ),
    M_age = mean(age, na.rm = TRUE),
    SD_age = sd(age, na.rm = TRUE),
    min_age = min(age, na.rm = TRUE),
    max_age = max(age, na.rm = TRUE),
    M_vviq = mean(vviq, na.rm = TRUE),
    SD_vviq = sd(vviq, na.rm = TRUE)
  ) |>
  dplyr::rename("VVIQ group" = 1) |>
  knitr::kable(digits = 2)
```

## Recruitment, and why the five studies aren't interchangeable {#recruitment-and-why}

These five studies were not designed together, and it shows in ways worth being
upfront about, not smoothing over. They differ in language (three
English-speaking samples, two French-speaking), in recruitment channel (social
media, a dedicated aphantasia research database, a university subject pool,
Prolific), and most consequentially, in how "aphantasia" was operationalised
before this project pooled everything onto a common 16-80 VVIQ scale. Ale &
Burns used a cutoff of VVIQ ≤ 32 to define their aphantasia group, explicitly
because their own prior work found that threshold more inclusive of people who
self-identify as having aphantasia than a stricter cutoff. Monzel et al. used
VVIQ ≤ 23. Kvamme et al. used ≤ 32 for their broader "aphantasia" group, then
further distinguished "core aphantasia" (16-23) from "hypophantasia" (24-32)
within it, a finer split closely related to the one this project's own
categorical model uses (see [model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)).

None of this is a flaw in any individual study, as each threshold was a reasonable
choice given that study's own goals. It is, however, a real illustration of a
point this project's own results speak to directly: VVIQ threshold conventions
vary across the field, sometimes substantially, and a pooled analysis that
treats VVIQ as continuous rather than pre-sorting everyone into study-specific
categories sidesteps that inconsistency rather than inheriting it.

## By dataset, VVIQ, and TAS-20 (alexithymia) group

```{r plot-proportions}
#| fig.alt: >
#|   Two combined panels. The top panel shows, for each of the five
#|   studies plus an aggregate 'total' bar, how many participants fall
#|   into each of the four VVIQ groups (aphantasia, hypophantasia,
#|   typical, hyperphantasia). The bottom panel shows the proportion of
#|   participants in each VVIQ group who score above the TAS-20's
#|   clinical alexithymia cutoff.

p_counts <-
  all_data |>
  dplyr::bind_rows(all_data |> dplyr::mutate(study = "total")) |>
  dplyr::mutate(
    study = factor(
      study,
      levels = c("burns", "monzel", "mas", "ruby", "kvamme", "total")
    )
  ) |>
  plot_vviq_group_proportions(vviq_group_4, base_size = 13, prop_txt_size = 5)

p_props <-
  all_data |>
  summarise_aph_and_alexi(vviq_group_4) |>
  plot_alexithymia_proportions(
    vviq_group_4,
    ncol = 6,
    base_size = 13,
    prop_txt_size = 3
  )

ggpubr::ggarrange(
  p_counts,
  p_props,
  ncol = 1,
  heights = c(1.1, 1),
  labels = "AUTO",
  font.label = list(size = 24, face = "bold")
) |> suppressWarnings()
```

The complete-aphantasia group (VVIQ = 16) is present in every study except Mas &
Luminet's, whose young, French, course-credit sample happened not to include
anyone at the scale's absolute floor, which is worth keeping in mind when reading the
[floor-group
model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html)
page's per-study breakdown, where that study's line is necessarily estimated
without any direct floor-group data of its own.

## Questionnaire reliability {#reliability}

Beyond the reliability figures reported for the original and French
validations of the VVIQ and TAS-20 (see the manuscript's *Questionnaires*
section), it is good practice to also report internal consistency for the
specific, pooled sample an analysis actually runs on: reliability is a
property of a sample's responses, not solely of an instrument in the
abstract, and pooling five studies collected under different conditions is the 
kind of situation where it is worth checking rather than assuming.

```{r reliability-setup}
items_flat <- 
  all_data |>
  dplyr::select(id, study, items) |>
  tidyr::unnest(items) |>
  dplyr::select(
    study, 
    dplyr::starts_with("vviq_q"), 
    dplyr::starts_with("tas_q"))
```

```{r reliability-omega}
#| results: hide

sample_omegas <-
  items_flat |>
  dplyr::select(!study) |> 
  tidyr::nest(
    vviq = dplyr::starts_with("vviq_q"),
    tas  = dplyr::starts_with("tas_q"),
    dif = c(tas_q1, tas_q3, tas_q6, tas_q7, tas_q9, tas_q13, tas_q14),
    ddf = c(tas_q2, tas_q4, tas_q11, tas_q12, tas_q17),
    eot = c(tas_q5, tas_q8, tas_q10, tas_q15, tas_q16, tas_q18, tas_q19, tas_q20)
  ) |>
  dplyr::rowwise() |> 
  dplyr::mutate(
    `VVIQ (16 items)` = psych::omega(vviq, plot = FALSE)$omega.tot,
    `TAS-20, total (20 items)` = psych::omega(tas, plot = FALSE)$omega.tot,
    `TAS-20, DIF (7 items)` = psych::omega(dif, plot = FALSE)$omega.tot,
    `TAS-20, DDF (5 items)` = psych::omega(ddf, plot = FALSE)$omega.tot,
    `TAS-20, EOT (8 items)` = psych::omega(eot, plot = FALSE)$omega.tot
  ) |> 
  dplyr::select(!c(vviq, tas, dif, ddf, eot)) |> 
  tidyr::pivot_longer(
    cols = tidyselect::everything(),
    names_to = "Scale",
    values_to = "McDonald's omega (total)"
  )
```

```{r reliability-display}
knitr::kable(sample_omegas, digits = 2)
```

```{r reliability-values}
#| echo: false

omega_vviq <- as.numeric(sample_omegas[1,2])
omega_tas <- as.numeric(sample_omegas[2,2])
omega_dif <- as.numeric(sample_omegas[3,2])
omega_ddf <- as.numeric(sample_omegas[4,2])
omega_eot <- as.numeric(sample_omegas[5,2])
```

In the present pooled sample (N = `r nrow(all_data)`), the VVIQ and the
TAS-20 total score both show excellent internal consistency (McDonald's
$\omega$ = `r round(omega_vviq, 2)` and
`r round(omega_tas, 2)`, respectively), comparable to or better
than the reliabilities reported for the original and French validations
cited in the manuscript. At the subscale level, DIF ($\omega$ =
`r round(omega_dif, 2)`) and DDF ($\omega$ =
`r round(omega_ddf, 2)`) are similarly strong, while EOT
($\omega$ = `r round(omega_eot, 2)`) is comparatively weaker,
consistent with the pattern already noted in the manuscript's
*Questionnaires* section, where EOT is described as the TAS-20's
least internally consistent facet across the wider literature
[@schroedersStructureTorontoAlexithymia2022; @bagby2020]. This is not a
given for a sample pooled across five studies, three languages of data
collection (English and French, see [Recruitment, and why the five
studies aren't
interchangeable](#recruitment-and-why-the-five-studies-arent-interchangeable)
above), and a wide range of recruitment channels: checking it directly,
rather than assuming it from the instruments' published psychometric
properties alone, is a small but genuinely useful piece of due diligence
that is not always done in practice.

We can also check McDonald's $\omega$ *per study*, except for @monzelAffectiveProcessingAphantasia2024, who did not provide item-level data in their open dataset:

```{r omega-per-study}
items_flat |>
  dplyr::group_by(study) |> 
  tidyr::nest(
    vviq = dplyr::starts_with("vviq_q"),
    tas  = dplyr::starts_with("tas_q"),
    dif = c(tas_q1, tas_q3, tas_q6, tas_q7, tas_q9, tas_q13, tas_q14),
    ddf = c(tas_q2, tas_q4, tas_q11, tas_q12, tas_q17),
    eot = c(tas_q5, tas_q8, tas_q10, tas_q15, tas_q16, tas_q18, tas_q19, tas_q20)
  ) |>
  dplyr::rowwise() |> 
  dplyr::mutate(
    omega_vviq = psych::omega(vviq, plot = FALSE)$omega.tot,
    omega_tas = psych::omega(tas, plot = FALSE)$omega.tot,
    omega_dif = psych::omega(dif, plot = FALSE)$omega.tot,
    omega_ddf = psych::omega(ddf, plot = FALSE)$omega.tot,
    omega_eot = psych::omega(eot, plot = FALSE)$omega.tot
  ) |> 
  dplyr::select(!c(vviq, tas, dif, ddf, eot)) |> 
  knitr::kable(digits = 2)
```


---

**Continuing through the Extended Online Report:** this page follows [how this
study found its
shape](https://m-delem.github.io/aphantasiaEmotions/articles/how-this-study-found-its-shape.html).
To keep reading in order, continue to the [model
comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
page next. Or jump to [the floor-group model, in
depth](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html),
[model diagnostics](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html),
[implementation notes](https://m-delem.github.io/aphantasiaEmotions/articles/implementation-notes.html),
or [for those who come
after](https://m-delem.github.io/aphantasiaEmotions/articles/for-those-who-come-after.html).

---

## References
