---
title: "Sample description"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sample description}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# NOTE: every figure-producing chunk needs a fig.alt option (pkgdown flags
# missing alt text; it is also good accessibility practice). Use Quarto-style
# #| pipe-comment chunk options throughout, as in how-this-study-found-its-shape.Rmd.

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
```

## Coming soon

This page will describe the five pooled datasets in detail — recruitment,
inclusion criteria, and sample composition per study — alongside the
distribution figures already built for this project.

---

**Continuing through the Extended Online Report:** this page follows
[how this study found its shape](https://m-delem.github.io/aphantasiaEmotions/articles/how-this-study-found-its-shape.html).
To keep reading in order, continue to the
[model comparison](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html)
page next. Or jump straight to the
[floor-group model](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html)
or [for those who come after](https://m-delem.github.io/aphantasiaEmotions/articles/for-those-who-come-after.html).
