---
title: "Superseded models"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Superseded models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 3.5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaEmotions)
library(ggplot2)
library(patchwork)

pkg   <- "aphantasiaEmotions"
refit <- "never"

options("marginaleffects_safe" = FALSE)
draws <- seq(1, 4000, 1) # To limit draws that will be used for marginaleffects
```

The 4-group categorical model and the Bayesian GAM were this study's
original, primary models, reported in [the first version of the manuscript](https://doi.org/10.31234/osf.io/es425_v1).
The [model-comparison arc](https://m-delem.github.io/aphantasiaEmotions/articles/model-comparison.html#comparing-all-single-level-models)
now prefers the floor-group and segmented models instead, as both fit the
data decisively better by LOO. This page is not a retraction of the
original models so much as a record of them: they are genuinely
informative, we checked them thoroughly, and we think it is worth showing
that check in full rather than letting them quietly disappear from the
documentation once something else took the spotlight.

The code and figures below are close to unchanged from the original,
single-vignette version of this package's report. They are reused here rather
than redone, since the underlying question they answer (how does each
model describe the data) hasn't changed, even though our headline answer
to the *overall* research question has moved on to a different model.

## Total TAS-20 scores

### 4-group categorical model {#original-4g}

```{r categorical-tot}
#| fig.alt: >
#|   Left: violin plots of total TAS score by VVIQ group (complete
#|   aphantasia, hypophantasia, typical imagery, hyperphantasia), with the
#|   clinical alexithymia cutoff marked. Right: posterior contrasts between
#|   each pair of groups, with ROPE-based evidence annotations.

lm_categorical_4g <- fit_brms_model(
  formula = tas ~ vviq_group_4,
  data    = all_data,
  prior   = brms::prior(normal(0, 20), class = "b"),
  file    = system.file("models", "lm_categorical_4g_tot.rds", package = pkg),
  file_refit = refit
)

contrasts_tot <- marginaleffects::comparisons(
  lm_categorical_4g,
  variables = list("vviq_group_4" = "pairwise"),
  draw_ids = draws
)

report_rope(contrasts_tot, contrast) |> knitr::kable(digits = 3)

p_contr_tot <- plot_posterior_contrasts(
  contrasts_tot,
  lm_categorical_4g,
  base_size = 12,
  rope_txt = 3,
  dot_size = 1,
  x_lab = "Effect size (TAS score difference)",
  axis_relative_x = 0.7
)

p_tot <- plot_group_violins(
  tas ~ vviq_group_4,
  y_lab = "Total TAS Score",
  base_size = 12
) +
  plot_alexithymia_cutoff(txt_size = 2, txt_x = 1.4, label = "Alexithymia") +
  scale_discrete_aphantasia() +
  scale_x_aphantasia(add = c(0.4, 0.7))

p_tot + p_contr_tot
```

(Convergence and posterior predictive check: [model diagnostics §Categorical, 4 groups](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html#ppc-categorical-4).)

### Bayesian GAM {#original-gam}

```{r gam-tot}
#| fig.alt: >
#|   Left: fitted GAM curve of total TAS score across the VVIQ range, with
#|   individual data points coloured by VVIQ group and the clinical
#|   alexithymia cutoff marked. Right: the estimated slope (rate of change
#|   of TAS with VVIQ) across the VVIQ range, coloured by evidence strength.

gam_tot <- fit_brms_model(
  formula = tas ~ s(vviq),
  data    = all_data,
  prior   = brms::prior(normal(0, 20), class = "b"),
  file    = system.file("models", "gam_tot.rds", package = pkg),
  file_refit = refit
)

slopes_tot <- modelbased::estimate_slopes(
  gam_tot,
  trend = "vviq",
  by = "vviq",
  length = 75,
  rope_ci = 1
)

check_slope_evidence(slopes_tot) |> knitr::kable(digits = 3)

p_slopes_tot <- plot_gam_slopes(
  slopes_tot,
  .f_groups = dplyr::case_when(
    vviq <= 24 ~ 1,
    vviq <= 35 ~ 2,
    vviq <= 36 ~ 3,
    vviq <= 45 ~ 4,
    vviq <= 76 ~ 5,
    vviq <= 80 ~ 6
  ),
  y_lab = "TAS variation per unit change in VVIQ",
  base_size = 12
)

p_gam_tot <- plot_gam_means(
  gam_tot,
  y_lab = "Total TAS score",
  legend_relative = 0.85,
  base_size = 12
) +
  plot_coloured_subjects(x = all_data$vviq, y = all_data$tas, size = 1) +
  plot_alexithymia_cutoff(txt_x = 26, label = "Alexithymia") +
  scale_discrete_aphantasia() +
  scale_x_vviq()

p_gam_tot + p_slopes_tot
```

(Convergence and posterior predictive check: [model diagnostics §GAM](https://m-delem.github.io/aphantasiaEmotions/articles/model-diagnostics.html#ppc-gam).)

## TAS-20 sub-scales

The same two models were fit separately on each of the three TAS-20 sub-scales
(DIF, DDF, EOT). Full figures and statistical resutls for these were part of the
original manuscript; they are not reproduced here since the
pattern across sub-scales did not depart dramatically from the one observed for
the total TAS. We refer interested readers to the [original manuscript](https://doi.org/10.31234/osf.io/es425_v1) for these details, or to 
the new final sub-scale analysis using the best model (the "floor-group" model) on
its [dedicated page](https://m-delem.github.io/aphantasiaEmotions/articles/floor-group-model.html#floor-subscales).
